---
output: github_document
---

```{r, include = FALSE}
options(readr.show_types = FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  fig.width = 6,
  fig.height = 4,
  dpi = 80,
  eval = TRUE
)
```

# Analysing the overlap between true and estimated segments

I like to use the package p_GenomicRanges_](https://bioconductor.org/packages/release/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesIntroduction.html) from Bioconductor for this type of analysis (checking overlaps between ranges, proportions of overlaps, etc.).

```{r}
# install.packages("BiocManager")
# BiocManager::install("GenomicRanges")

library(GenomicRanges)
library(tidyverse)
library(glue)
```


Load the true segments:

```{r}
true_segments <-
  read_tsv("~/Cluster/introgression-sims/results/tracts.tsv.gz",
           show_col_types = FALSE) %>%
  mutate(chrom = "chr1") %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
```

Get a basic summary of the data:

```{r}
summary(width(true_segments))
```

```{r, ind_distributions}
as_tibble(true_segments) %>%
  ggplot() +
  geom_density(aes(width / 1e3, color = factor(slim_id)), alpha = 0.5) +
  geom_density(aes(width / 1e3)) +
  geom_vline(xintercept = mean(width(true_segments) / 1e3), linetype = 2, alpha = 0.75) +
  geom_vline(xintercept = median(width(true_segments) / 1e3), linetype = 3, alpha = 0.75) +
  guides(color = "none") +
  xlab("length of a true segment [kb]") +
  coord_cartesian(xlim = c(0, 500)) +
  ggtitle("Length distribution of true Neanderthal segments (simulated)",
          glue("dotted line: median length [{format(median(width(true_segments) / 1e3),",
               "digits = 1)} kb], dashed line: mean length [{format(mean(width(true_segments) / 1e3), digits = 2)} kb"))
```

Create fake "estimated" date by messing up the known true segments (in practice these would be the fragments estimated by the software):

```{r}
filtered_segments <- true_segments[width(true_segments) > 20e3]
filtered_segments$set <- "true"

# create fake segments which extend by 10kb on both ends of the true segments
est_segments_large <- filtered_segments
start(est_segments_large) <- start(est_segments_large) - 10000
end(est_segments_large) <- end(est_segments_large) + 10000
est_segments_large$set <- "large"

# create fake segments which are 10kb shorter on both ends of the true segments
est_segments_small <- filtered_segments
start(est_segments_small) <- start(est_segments_small) + 10000
end(est_segments_small) <- end(est_segments_small) - 10000
est_segments_small$set <- "small"

# create fake segments which are shifted by 25% of the length
est_segments_shifted <- filtered_segments
start(est_segments_shifted) <- start(est_segments_shifted) + width(est_segments_shifted) / 4
est_segments_shifted$set <- "shifted"
```

```{r, fake_distributions}
# plot the overall distributions of all three sets of segments

segments <- rbind(
  as_tibble(filtered_segments),
  as_tibble(est_segments_large),
  as_tibble(est_segments_small),
  as_tibble(est_segments_shifted)
)

segments %>%
  ggplot() +
  geom_density(aes(width / 1e3, color = set), alpha = 0.5) +
  xlab("length of a segment [kb]") +
  coord_cartesian(xlim = c(0, 500)) +
  ggtitle("Length distribution of true Neanderthal segments (simulated)")
```

Now let's try to "analyse" the overlap of the "estimated" segments with the true segments for one particular chromosome:

```{r}
slim_id <- 476018414
```

### Sanity check #1 - overlap of segments with themselves

This must give overlap of 100%:

```{r, overlap_true_with_true}
query <- filtered_segments
subject <- filtered_segments

query <- query[query$slim_id == slim_id]
subject <- subject[subject$slim_id == slim_id]

hits <- findOverlaps(query, subject)

overlaps <- pintersect(query[queryHits(hits)], subject[subjectHits(hits)])
proportions <- width(overlaps) / width(subject[subjectHits(hits)])

qplot(proportions) + geom_vline(xintercept = c(0, 1)) + xlab("overlap proportion")
```

### Overlap of segments with those that are smaller

This must be less than 100% by definition:

```{r, overlap_true_with_small}
query <- est_segments_small
subject <- filtered_segments

query <- query[query$slim_id == slim_id]
subject <- subject[subject$slim_id == slim_id]

hits <- findOverlaps(query, subject)

overlaps <- pintersect(query[queryHits(hits)], subject[subjectHits(hits)])
proportions <- width(overlaps) / width(subject[subjectHits(hits)])

qplot(proportions) + geom_vline(xintercept = c(0, 1)) + xlab("overlap proportion")
```


### Overlap of segments with those that are larger

Has to be 100% by definition because we have "estimated" outside of each true fragment:

```{r, overlap_true_with_large}
query <- est_segments_large
subject <- filtered_segments

query <- query[query$slim_id == slim_id]
subject <- subject[subject$slim_id == slim_id]

hits <- findOverlaps(query, subject)

overlaps <- pintersect(query[queryHits(hits)], subject[subjectHits(hits)])
proportions <- width(overlaps) / width(subject[subjectHits(hits)])

qplot(proportions) + geom_vline(xintercept = c(0, 1)) + xlab("overlap proportion")
```


### Overlap of segments with those that are shifted

This has to be ~75% by definition because above we shifted the start of each segment by 25% of its length:

```{r, overlap_true_with_shifted}
query <- est_segments_shifted
subject <- filtered_segments

query <- query[query$slim_id == slim_id]
subject <- subject[subject$slim_id == slim_id]

hits <- findOverlaps(query, subject)

overlaps <- pintersect(query[queryHits(hits)], subject[subjectHits(hits)])
proportions <- width(overlaps) / width(subject[subjectHits(hits)])

qplot(proportions) + geom_vline(xintercept = c(0, 1)) + xlab("overlap proportion")
```
